Binning Strategies and Related Loss for Binned Scatterplots of Large Data

### 2016-10-04 versionForJCGS/PaperJCGS.Rnw
minor changes

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}
\newcommand{\ktm}[1]{{\color{red} #1}} %red comments: Karsten
\newcommand{\hh}[1]{{\color{blue} #1}} %blue comments: Heike

@@

<<setup4,echo=F,include=F,eval=T>>=

# library(dbData)
library(devtools)
library(ggplot2)
library(gridExtra)
library(hexbin)
library(Hmisc)
library(lubridate)
library(MASS)
library(plyr)
library(RColorBrewer)
library(reshape2)
library(scales)
library(tidyr)
library(xtable)
library(scagnostics)

#source in binning functions
source("BinningLossFunctions.R")

@@

# Load all data: pitching and binning loss demo
load("./data/pitching.Rdata")
load("./data/BinningLossDemos.Rdata")
@

@@

# # Hex binned scatterplot
#   qplot(data=pitchdat, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
#   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
#   ggsave("./figure/HexBinning.png", width=5, height=5, dpi=300)
d1.hex.obj <- hexbin(pitchdat$G, pitchdat$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <-  d1.hex.obj@count

@@

<<DataReduction,echo=FALSE,eval=FALSE,results='asis'>>=
  # Generate tables for raw, binned and reduced binned data
n=12
param1 = 1
param2 = 1.6
set.seed(52)
x <- rbeta(n,param1, param2)*20-10
y <- rbeta(n,param1, param2)*20-10
# standard bin centers
xcent <- rep(-5,n) ; xcent[x>0] <- 5
ycent <- rep(-5,n) ; ycent[y>0] <- 5
idx <- order(xcent,ycent)
d <- data.frame(x,y,xcent,ycent)[idx,]
dbin <- ddply(d, .(xcent,ycent), summarise,
              c=length(x))
dbin <- dbin[order(dbin$xcent,dbin$ycent),]

@@

#match color scales in each
low = "#132B43"
high = "#56B1F7"

@@

qplot(x, y, colour=1000*Shade, size=I(5),shape=Shape) + theme_bw() + 
  xlab("") + ylab("")+
  theme(plot.margin = unit(c(0.1,0.1,0.1,.1),"cm"), aspect.ratio=1,
        axis.ticks=element_blank(), axis.text=element_blank()) +
  scale_colour_gradient("Shade",limits=c(0,1000), breaks=c(0,500,1000), 
                        low=high, high=low)

@@

low = "#132B43"
high = "#56B1F7"
blueGradient <- matrix(seq_gradient_pal(high, low, "Lab")(0:4400/4400), byrow=TRUE,
       nrow=1000, ncol=4401)
g <- grid::rasterGrob(blueGradient, interpolate=TRUE)

@@

Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with that binning algorithm. This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of distributional scenarios. Some choices may be proven optimal through analytical properties, while other are data dependent and require empirical exploration of loss from binning. Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.

Data sets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform. 100,000 observation pairs were simulated for each data set from the following distributions:

@@

\item Set III: $x_i \sim \text{iid Uniform}(a,b), y_i \sim \text{iid Uniform}(a,b)$ with $a =0$ and $b=100$
\end{itemize}
%
The parameters were selected to have data values roughly span the region $[0,100]^2$. The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}. These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places; 6 decimal places in our simulate data. Real data is recorded to only the number of digits that measurement precision allows, and in many cases rounded even further.

%
<<SimDataScatterplots2,echo=F,include=F,eval=T,fig.width=8, fig.height=5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data. Density changes are impossible to make out in the solid black areas of the plots. 97.4\\%, 98.7\\% and 99.1\\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.">>=

### Try 2d binning with MVN data
n=100000
set.seed(31415)
unifdat <- data.frame(x= runif(n, min=0, max=100),
                      y= runif(n, min=0, max=100),
                      type="Uniform",
                      record = "Fine")
normdat <- data.frame(x= rnorm(n, mean=50, sd=11),
                     y= rnorm(n, mean=50, sd=11),
                     type="Normal",
                     record = "Fine")
expdat <- data.frame(x= rexp(n, .11),
                     y= rexp(n, .11),
                     type="Exponential",
                     record = "Fine")
datlist <- list(unifdat,normdat,expdat)
### round all data to resolution alpha
alpha <- 2
for(i in 1:3){
  datlist[[i]]$roundx <- ((datlist[[i]]$x + 1)%/%alpha)*alpha
  datlist[[i]]$roundy <- ((datlist[[i]]$y + 1)%/%alpha)*alpha
}

@@

ggsave("./figure/SimScatters.png", width=8, height=5.5, dpi=300)
@
\begin{figure}[hbtp]
\begin{center}
 \includegraphics[keepaspectratio=TRUE,width=.9\textwidth]{./figure/SimScatters.png}
	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data. }
	\end{center}
\end{figure}
%
%Density changes are impossible to make out in the solid black areas of the plots. 97.4\%, 98.7\% and 99.1\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.
%
The \textit{data resolution} is defined as the smallest increment between successive data values. To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, hus a data resolution of 2 units in each dimension. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.

\ktm{These bivariate distributions were selected for the variety in their defining characteristics. \textit{Scagnostics} - a portmanteau for scatterplot diagnostics - are formalized metrics that have been used to numerically summarize key attributes of scatterplots, including: outying, skewness, clumpiness, sparsity, striation, convexity, skinniness, stringiness and monotonicity (\citealt{wilkinson2005graph};\citealt{wilkinson2008scagnostics}). The scagnostic scores in Table~\ref{tab:scagnostics} were calculated for the simulated data sets using the \texttt{scagnostics} package in R \citep{wilkinson2015package}. They show that these examples explore a variety of skewnesses, convexity and skinniness. This exploration uses bivariately independent distributions, thus the monotonicity and clumpiness are both near zero for these examples. The coarse data is rounded into striatations by definition, however the striation scagnostic score is low due to the relatively small scale of striation versus the magnitude of variability in each distribution. These minor striations will still prove problematic when binning is inconsistent with the data resolution. }

<<SimDataScagnostics,echo=F,include=F,eval=F>>= 
# help(package="scagnostics")
head(alldat)
scags <- data.frame(0,0,0,0,0,0,0,0,0,0,0)
counter <- 0
for (i in c("Exponential","Normal","Uniform")){
  for(j in c("Fine","Coarse")){
    counter <- counter + 1
    tmp <- filter(alldat, type==i, record=="Fine")
    print(dim(tmp))
    if(j=="Fine") scags[counter,] <- c(i,j,round(as.vector(scagnostics(tmp$x,tmp$y)),2))
    if(j=="Coarse") scags[counter,] <- c(i,j,round(as.vector(scagnostics(tmp$roundx,tmp$roundy)),2))   
    names(scags) <- c("type","record",names(scagnostics(tmp$x,tmp$y)))
  }
}
head(scags)

print(xtable(scags, digits=c(0, 0, 3,3,3,3,3,3,3,3,3,3,3)), include.rownames=FALSE)
@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Sat Oct 01 11:29:18 2016
\begin{table}[ht]
\ktm{
\centering \small
\begin{tabular}{lrrrrrrrrr}
  \hline
Dist-Res & Outlie & Skew & Clump & Sparse & Striate & Convex & Skinny & String & Mono. \\ 
  \hline
  Exp-Fine &   0.15 & 0.67 & 0.01 & 0.02 & 0.02 & 0.47 & 0.55 & 0.36 & 0.00 \\ 
  Exp-Coarse &  0.15 & 0.71 & 0.02 & 0.02 & 0.03 & 0.51 & 0.52 & 0.27  & 0.00  \\ 
  Norm-Fine &   0.07 & 0.55 & 0.00 & 0.02 & 0.01 & 0.60 & 0.31 & 0.39 & 0.00  \\ 
  Norm-Coarse &   0.08 & 0.59 & 0.01 & 0.02 & 0.02 & 0.60 & 0.33 & 0.35  & 0.00  \\ 
  Unif-Fine &   0.00 & 0.52 & 0.00 & 0.03 & 0.00 & 0.68 & 0.15 & 0.36 & 0.00  \\ 
  Unif-Coarse & 0.00 & 0.56 & 0.00 & 0.03 & 0.01 & 0.68 & 0.14 & 0.38  & 0.00  \\ 
   \hline
\end{tabular}
} % end ktm highlighting
\normalsize
\caption{\ktm{Scagnostics scores for the six simulated datasets used in empirical exploration of spatial loss in binned scatterplots.}}
\label{tab:scagnostics}
\end{table}

@@

If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency. Binned scatterplots shift visual emphasis from true location of individual points to the geometric centers of tiles, making them visually biased displays of density. However, as bin sizes become increasingly fine, the bias decreases; with no visual bias when minimal binning perfectly matches bins with the data resolution the density estimate. Also, the density estimation more perfectly reflect the bivariate density as the sample size increases; thus making the binned scatterplot a consistent visual estimator. We may also consider a minimal binned scatterplot as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator. It is worth noting that a density estimate requires the combination of spatial and frequency information and that the estimation properties were considered only through the lens of spatial loss. This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies. %Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.  
%
<<SpatialLossCompare,echo=F,warning=FALSE,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss and computation times over a range of bin sizes for standard and random binning.">>= 
## create binned data for standard and random using variety of bin sizes (from 2 to 20, by 2)
##save spatial loss information
# spatlossdat <- data.frame(NULL)
# index=0
# distindex = 0
# distrib <- c("Uniform", "Normal", "Exponential")
# for (dat in 1:3){
#   distindex <- distindex + 1
#   for (width in seq(2,20,by=2)){
#       index <- index+1
#       timeStart <- Sys.time()
#       binout <- RectBin2d(datlist[[dat]]$x,datlist[[dat]]$y,0,0,width,width,type="standard")
#       binout[[2]]$binTime <- Sys.time() - timeStart
#       binout[[2]]$bintype <- "standard"
#       binout[[2]]$distrib <- distrib[distindex]
#       spatlossdat <- rbind(spatlossdat, binout[[2]])
#       index <- index+1
#       timeStart <- Sys.time()
#       binout <- RectBin2d(datlist[[dat]]$x,datlist[[dat]]$y,0,0,width,width,type="random")
#       binout[[2]]$binTime <- Sys.time() - timeStart
#       binout[[2]]$bintype <- "random"
#       binout[[2]]$distrib <- distrib[distindex]
#       spatlossdat <- rbind(spatlossdat, binout[[2]])
#       print(c(distrib[distindex],width))
#   }
# }
# # write.csv(spatlossdat,"SpatialLossData.csv", row.names=FALSE)
# spatlossdatSquareBins <- read.csv("SpatialLossData.csv",header=TRUE) 
# NOTE: spatlossdatSquareBins saved to BinningLossDemos.Rdata which is read in chunk 1
spatlossdatSquareBins$bintype <- factor(spatlossdatSquareBins$bintype, levels=c("standard","random"))

spatlossdatSquareBins2 <- gather(spatlossdatSquareBins, measure, value, c(totalSpatialLoss,binTime))
spatlossdatSquareBins2$measure <- factor(spatlossdatSquareBins2$measure , labels=c("Binning Time (sec)","Net Spatial Loss"))
# 
# p1 <- qplot(widthx,totalSpatialLoss,geom="path",colour=bintype, linetype=bintype,
#       data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
#   theme_bw() + ylab("Net Spatial Loss") +
#   xlab("Width/Height of Square Bins") + ylim(c(0, 1100000))
# 
# p2 <- qplot(widthx,as.numeric(binTime),geom="path",colour=bintype, linetype=bintype,
#       data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
#   theme_bw() + ylab("Binning Time (sec)") +
#   xlab("Width/Height of Square Bins")
# 
# grid.arrange(p1,p2, nrow=2)

qplot(widthx,value,geom="path",colour=bintype, linetype=bintype,
      data=spatlossdatSquareBins2,group=bintype, size=I(1)) +
  facet_grid(measure~distrib, scales="free_y" )+
  theme_bw() + theme(plot.margin = unit(c(0.1,0.1,0.1, 0.1),"cm")) + 
  ylab("Net Spatial Loss") +
  xlab("Size of Bin Dimensions") 
@
%
%All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin. 
While using smaller bin widths leads to smaller spatial losses, reducing bin sizes comes at the cost of computation time; a potentially non-negligible consideration in settings with truly massive data. Figure~\ref{fig:SpatialLossCompare} also shows that the computation time needed to bin the simulated data sets is a decreasing function of the bin size and that random binning is marginally slower than standard binning across all bin sizes.\footnote{Computation times using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz.} Larger bins may also be reasonable to use if we are primary interested visualizing the large scale density structure and wish to smooth over fine structural noise.

@@

<<UniformStripes, echo=F,include=T,eval=T,fig.width=7, fig.height=3, out.width='.99\\linewidth', fig.pos='hbtp',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots of coarse uniform data.">>= 
# Commented out code for data generation: run once then recomment
### Create data to demonstrate stripes from improper binwidths
# binoutcoarse1 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,1,1,type="standard")[[1]]
# binoutcoarse4 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,4,4,type="standard")[[1]]
# binoutcoarse5 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="standard")[[1]]
# binoutcoarse1Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,1,1,type="random")[[1]]
# binoutcoarse4Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,4,4,type="random")[[1]]
# binoutcoarse5Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="random")[[1]]
# 
# binoutcoarse1$BinSize <- "1 X 1 Unit" ; binoutcoarse1$BinningType <- "Standard" 
# binoutcoarse4$BinSize <- "4 X 4 Unit" ; binoutcoarse4$BinningType <- "Standard" 
# binoutcoarse5$BinSize <- "5 X 5 Unit" ; binoutcoarse5$BinningType <- "Standard" 
# binoutcoarse1Rand$BinSize <- "1 X 1 Unit" ; binoutcoarse1Rand$BinningType <- "Random" 
# binoutcoarse4Rand$BinSize <- "4 X 4 Unit" ; binoutcoarse4Rand$BinningType <- "Random" 
# binoutcoarse5Rand$BinSize <- "5 X 5 Unit" ; binoutcoarse5Rand$BinningType <- "Random" 
# 
# allbindat <- rbind(binoutcoarse1,binoutcoarse4,binoutcoarse5,
#                    binoutcoarse1Rand,binoutcoarse4Rand,binoutcoarse5Rand)
# write.csv(allbindat,"StripeBinData.csv", row.names=FALSE)
# allbindat <- read.csv("StripeBinData.csv",header=TRUE)
# NOTE: allbindat saved to BinningLossDemos.Rdata which is read in chunk 1

allbindat$BinningType <- factor(allbindat$BinningType, levels=c("Standard","Random"))
allbindat$height <- 1 ; allbindat$width <- 1
allbindat$height[which(allbindat$BinSize=="4 X 4 Unit")] <- 4
allbindat$width[which(allbindat$BinSize=="4 X 4 Unit")] <- 4
allbindat$height[which(allbindat$BinSize=="5 X 5 Unit")] <- 5
allbindat$width[which(allbindat$BinSize=="5 X 5 Unit")] <- 5

@@

<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=7, out.width='.99\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively.",fig.scap="Lineplots for total frequency losses from standard and quantile binning for simulated bivariate data. ">>= 
### Loop over ncolors 1 to 10 and also over data distribution
# find freq loss from standard, quantile, logCount
freqLossSimDat <- NULL
distrib <- c("Uniform", "Normal", "Exponential")
for (distidx in 1:3){
  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
  for (ncol in 1:10){
    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]], 
                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=TRUE)[[2]])
    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Standard Log", "Quantile Log")
    binoutFreq$DistType <- distrib[distidx]
    binoutFreq$ncolor <- ncol 
    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
  }
}

@@

%' THE FOLLOWING SUBSECTION IS OMITTED FOR JCGS PAGE LIMIT CONSIDERATIONS

>>>>

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}
\newcommand{\ktm}[1]{{\color{red} #1}} %red comments: Karsten
\newcommand{\hh}[1]{{\color{blue} #1}} %blue comments: Heike

@@

<<setup4,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
library(reshape2)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)
library(hexbin)
library(MASS)
library(scales)
library(plyr)
#source in binning functions
source("BinningLossFunctions.R")

@@

load("./data/pitching.Rdata")
@

@@

# Hex binned scatterplot
ggplot(data=pitchdat, aes(x=G, y=SO))+
    geom_hex()+
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
  ggsave("./figure/HexBinning.png", width=5, height=5, dpi=300)
d1.hex.obj <- hexbin(pitchdat$G, pitchdat$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <-  d1.hex.obj@count

@@

<<DataReduction,echo=FALSE,eval=FALSE,results='asis'>>=
n=12
param1 = 1
param2 = 1.6
set.seed(52)
x <- rbeta(n,param1, param2)*20-10
y <- rbeta(n,param1, param2)*20-10
# standard bin centers
xcent <- rep(-5,n) ; xcent[x>0] <- 5
ycent <- rep(-5,n) ; ycent[y>0] <- 5
idx <- order(xcent,ycent)
d <- data.frame(x,y,xcent,ycent)[idx,]
dbin <- ddply(d, .(xcent,ycent), summarise,
              c=length(x))
dbin <- dbin[order(dbin$xcent,dbin$ycent),]

@@

library(scales)
#match color scales in each
low = "#132B43"
high = "#56B1F7"

@@

require(ggplot2)
qplot(x, y, colour=1000*Shade, size=I(5),shape=Shape) + theme_bw() + 
  xlab("") + ylab("")+
  theme(plot.margin = unit(c(0.1,0.1,0.1,.1),"cm"), aspect.ratio=1,
        axis.ticks=element_blank(), axis.text=element_blank()) +
  scale_colour_gradient("Shade",limits=c(0,1000), breaks=c(0,500,1000), 
                        low=high, high=low)

@@

low = "#132B43"
high = "#56B1F7"
blueGradient <- matrix(seq_gradient_pal(high, low, "Lab")(0:4400/4400), byrow=TRUE,
       nrow=1000, ncol=4401)
g <- rasterGrob(blueGradient, interpolate=TRUE)

@@

Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with that binning algorithm. This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of scenarios. Some choices may be proven optimal through analytical properties, while other are data dependent and require empirical exploration of loss from binning.  Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.

Data sets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.  These distributions were selected for their variety in terms of shape, center and spread. 100,000 observation pairs were simulated for each data set from the following distributions:

@@

\item Set III: $x_i \sim \text{iid Uniform}(a,b), y_i \sim \text{iid Uniform}(a,b)$ with $a =0$ and $b=100$
\end{itemize}
%
The parameters were selected to have data values roughly span the region $[0,100]^2$. The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}. These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places; 6 decimal places in our simulate data. Real data is recorded to only the number of digits that measurement precision allows, and in many cases rounded even further.
%
<<SimDataScatterplots2,echo=F,include=F,eval=T,fig.width=8, fig.height=5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data. Density changes are impossible to make out in the solid black areas of the plots. 97.4\\%, 98.7\\% and 99.1\\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.">>= 
library(scales) 
### Try 2d binning with MVN data
n=100000
set.seed(31415)
unifdat <- data.frame(x= runif(n, min=0, max=100),
                      y= runif(n, min=0, max=100),
                      type="Uniform",
                      record = "Fine")
normdat <- data.frame(x= rnorm(n, mean=50, sd=11),
                     y= rnorm(n, mean=50, sd=11),
                     type="Normal",
                     record = "Fine")
expdat <- data.frame(x= rexp(n, .11),
                     y= rexp(n, .11),
                     type="Exponential",
                     record = "Fine")
datlist <- list(unifdat,normdat,expdat)
### round all data to resolution alpha
alpha <- 2
for(i in 1:3){
  datlist[[i]]$roundx <- ((datlist[[i]]$x + 1)%/%alpha)*alpha
  datlist[[i]]$roundy <- ((datlist[[i]]$y + 1)%/%alpha)*alpha
}

@@

ggsave("./figure/SimScatters.png", width=8, height=5.5, dpi=300)
@
\begin{figure}[hbtp]
\begin{center}
 \includegraphics[keepaspectratio=TRUE,width=.9\textwidth]{./figure/SimScatters.png}
	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data. }
	\end{center}
\end{figure}
%
%Density changes are impossible to make out in the solid black areas of the plots. 97.4\%, 98.7\% and 99.1\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.
%
The \textit{data resolution} is defined as the smallest increment between successive data values. To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units in each dimension. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.

@@

If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency. Binned scatterplots shift visual emphasis from true location of individual points to the geometric centers of tiles, making them visually biased displays of density. However, as bin sizes become increasingly fine, the bias decreases; with no visual bias when minimal binning perfectly matches bins with the data resolution the density estimate. Also, the density estimation more perfectly reflect the bivariate density as the sample size increases; thus making the binned scatterplot a consistent visual estimator. We may also consider a minimal binned scatterplot as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator. It is worth noting that a density estimate requires the combination of spatial and frequency information and that the estimation properties were considered only through the lens of spatial loss. This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies. %Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.  
%
<<SpatialLossCompare,echo=F,warning=FALSE,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss and computation times over a range of bin sizes for standard and random binning.">>= 
## create binned data for standard and random using variety of bin sizes (from 2 to 20, by 2)
##save spatial loss information
# spatlossdat <- data.frame(NULL)
# index=0
# distindex = 0
# distrib <- c("Uniform", "Normal", "Exponential")
# for (dat in 1:3){
#   distindex <- distindex + 1
#   for (width in seq(2,20,by=2)){
#       index <- index+1
#       timeStart <- Sys.time()
#       binout <- RectBin2d(datlist[[dat]]$x,datlist[[dat]]$y,0,0,width,width,type="standard")
#       binout[[2]]$binTime <- Sys.time() - timeStart
#       binout[[2]]$bintype <- "standard"
#       binout[[2]]$distrib <- distrib[distindex]
#       spatlossdat <- rbind(spatlossdat, binout[[2]])
#       index <- index+1
#       timeStart <- Sys.time()
#       binout <- RectBin2d(datlist[[dat]]$x,datlist[[dat]]$y,0,0,width,width,type="random")
#       binout[[2]]$binTime <- Sys.time() - timeStart
#       binout[[2]]$bintype <- "random"
#       binout[[2]]$distrib <- distrib[distindex]
#       spatlossdat <- rbind(spatlossdat, binout[[2]])
#       print(c(distrib[distindex],width))
#   }
# }
# # write.csv(spatlossdat,"SpatialLossData.csv", row.names=FALSE)
spatlossdatSquareBins <- read.csv("SpatialLossData.csv",header=TRUE)
spatlossdatSquareBins$bintype <- factor(spatlossdatSquareBins$bintype, levels=c("standard","random"))

library(tidyr)
spatlossdatSquareBins2 <- gather(spatlossdatSquareBins, measure, value, c(totalSpatialLoss,binTime))
spatlossdatSquareBins2$measure <- factor(spatlossdatSquareBins2$measure , labels=c("Net Spatial Loss","Binning Time (sec)"))
# 
# p1 <- qplot(widthx,totalSpatialLoss,geom="path",colour=bintype, linetype=bintype,
#       data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
#   theme_bw() + ylab("Net Spatial Loss") +
#   xlab("Width/Height of Square Bins") + ylim(c(0, 1100000))
# 
# p2 <- qplot(widthx,as.numeric(binTime),geom="path",colour=bintype, linetype=bintype,
#       data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
#   theme_bw() + ylab("Binning Time (sec)") +
#   xlab("Width/Height of Square Bins")
# 
# grid.arrange(p1,p2, nrow=2)

qplot(widthx,value,geom="path",colour=bintype, linetype=bintype,
      data=spatlossdatSquareBins2,group=bintype, size=I(1)) +
  facet_grid(measure~distrib, scales="free_y" )+
  theme_bw() + theme(plot.margin = unit(c(0.1,0.1,0.1, 0.1),"cm")) + 
  ylab("Net Spatial Loss") +
  xlab("Width/Height of Square Bins") 
@
%
%All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin. 
While using smaller bin widths leads to smaller spatial losses, reducing bin sizes comes at the cost of computation time; a potentially non-negligible consideration in settings with truly massive data. Figure~\ref{fig:SpatialLossCompare} also shows that the computation time needed to bin the simulated data sets is a decreasing function of the bin size and that random binning is marginally slower than standard binning across all bin sizes.\footnote{Computation times using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz.} Larger bins may also be reasonable to use if we are primary interested visualizing the large scale density structure and wish to smooth over fine structural noise.

@@

<<UniformStripes, echo=F,include=T,eval=T,fig.width=7, fig.height=3, out.width='.99\\linewidth', fig.pos='hbtp',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots of coarse uniform data.">>= 
# Commented out code for data generation: run once then recomment
### Create data to demonstrate stripes from improper binwidths
# binoutcoarse1 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,1,1,type="standard")[[1]]
# binoutcoarse4 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,4,4,type="standard")[[1]]
# binoutcoarse5 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="standard")[[1]]
# binoutcoarse1Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,1,1,type="random")[[1]]
# binoutcoarse4Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,4,4,type="random")[[1]]
# binoutcoarse5Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="random")[[1]]
# 
# binoutcoarse1$BinSize <- "1 X 1 Unit" ; binoutcoarse1$BinningType <- "Standard" 
# binoutcoarse4$BinSize <- "4 X 4 Unit" ; binoutcoarse4$BinningType <- "Standard" 
# binoutcoarse5$BinSize <- "5 X 5 Unit" ; binoutcoarse5$BinningType <- "Standard" 
# binoutcoarse1Rand$BinSize <- "1 X 1 Unit" ; binoutcoarse1Rand$BinningType <- "Random" 
# binoutcoarse4Rand$BinSize <- "4 X 4 Unit" ; binoutcoarse4Rand$BinningType <- "Random" 
# binoutcoarse5Rand$BinSize <- "5 X 5 Unit" ; binoutcoarse5Rand$BinningType <- "Random" 
# 
# allbindat <- rbind(binoutcoarse1,binoutcoarse4,binoutcoarse5,
#                    binoutcoarse1Rand,binoutcoarse4Rand,binoutcoarse5Rand)
# write.csv(allbindat,"StripeBinData.csv", row.names=FALSE)

allbindat <- read.csv("StripeBinData.csv",header=TRUE)
allbindat$BinningType <- factor(allbindat$BinningType, levels=c("Standard","Random"))
allbindat$height <- 1 ; allbindat$width <- 1
allbindat$height[which(allbindat$BinSize=="4 X 4 Unit")] <- 4
allbindat$width[which(allbindat$BinSize=="4 X 4 Unit")] <- 4
allbindat$height[which(allbindat$BinSize=="5 X 5 Unit")] <- 5
allbindat$width[which(allbindat$BinSize=="5 X 5 Unit")] <- 5

@@

<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=7, out.width='.99\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively.",fig.scap="Lineplots for total frequency losses from standard and quantile binning for simulated bivariate data. ">>= 
### Loop over ncolors 1 to 10 and also over data distribution
# find freq loss from standard, quantile, logCount
source("BinningLossFunctions.R")
freqLossSimDat <- NULL
distrib <- c("Uniform", "Normal", "Exponential")
for (distidx in 1:3){
  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
  for (ncol in 1:10){
    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]], 
                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=TRUE)[[2]])
    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Standard Log", "Quantile Log")
    binoutFreq$DistType <- distrib[distidx]
    binoutFreq$ncolor <- ncol 
    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
  }
}

@@
